\textbf{Towards transparency of hybrid open access through publisher-provided metadata: An article-level study of Elsevier}

### 2021-05-19 analysis/manuscript.tex
align text

\title{\textbf{Transparency to hybrid open access  through publisher-provided metadata: An article-level study of Elsevier}}
\shorttitle{Hybrid OA Elsevier}
\threeauthors{Najko Jahn}{Lisa Matthias}{Mikael Laakso}
\threeaffiliations{Göttingen State and University Library, University of Göttingen\\
Platz der Göttinger Sieben 1, 37073 Göttingen, Germany\\
najko.jahn@sub.uni-goettingen.de
}{Department of Political Science, John F. Kennedy Institute, Freie Universität Berlin \\
Lansstraße 7-9, 14195 Berlin, Germany \\
l.a.matthia@gmail.com
}{Information Systems Science, Hanken School of Economics \\
Arkadiankatu 22, 00100 Helsinki, Finland \\
mikael.laakso@hanken.fi
}

@@

\abstract{With the growth of open access (OA), the financial flows in scholarly journal
publishing have become increasingly complex, but comprehensive data and
transparency into these flows are still lacking. The opaqueness is especially
concerning for hybrid OA, where subscription-based journals publish individual
articles as OA if an optional fee is paid. This study addresses the lack of
transparency by leveraging Elsevier article metadata and provides the first
longitudinal publisher-level study of hybrid OA uptake and invoicing. Our
results show that Elsevier's hybrid OA uptake has grown steadily but slowly from
2015-2019, doubling the number of hybrid OA articles published per year and
increasing the share of OA articles in Elsevier's hybrid journals from 2.6\% to
3.7\% of all articles. Further, we find that most hybrid OA articles were
invoiced directly to authors, followed by articles invoiced through agreements
with research funders, institutions, or consortia, with only a few funding
bodies driving hybrid OA uptake. As such, our findings point to the role of
publishing agreements and OA policies in hybrid OA publishing. Our results
further demonstrate the value of publisher-provided metadata improve the
transparency in scholarly publishing by linking invoicing data to bibliometrics.}

@@

The rise of open access (OA) to journal articles has added complexity to
scholarly publishing, particularly concerning transparency of economic
dimensions. Financial transparency in journal publishing has long been
lacking since Big Deal subscription contracts between academic
institutions and large academic publishers usually include
confidentiality clauses that prohibit publicly sharing information about
the scope and pricing of such agreements (Bergstrom et al., 2014;
Frazier, 2001; Larivière et al., 2015). Further, the distributed process
of paying for accessing and publishing scholarly articles, which can
involve several actors at the same time, including authors, research
organizations, libraries, and research funders, adds to the opacity of
the financial flows of scholarly publishing (Lawson et al., 2016).

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available for a fee, also known as article-processing charge
(APC), while the journal as a whole remains subscription access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases (D. C. Prosser, 2003).
Many large subscription publishers have since incorporated this OA
model, and while the uptake among authors increased only slowly during
the first decade (Björk, 2012), it has gradually gained popularity in
recent years. This development was driven by the implementation of OA
policies by research institutions and funders, the allocation of OA
funds, and formal agreements between publishers and institutions that
allow affiliated authors to publish OA free-of-charge (Laakso \& Björk,
2016). However, there have been concerns that the hybrid model allows
for double-dipping---receiving two payments for one article, the APC and
subscription fees (D. Prosser, 2015; Shieber, 2009). While publishers
assure they combine subscription fees and APCs to adjust pricing
(Mittermaier, 2015), without transparency around the uptake of hybrid OA
and revenue streams, such claims are impossible to evaluate.

With the recent introduction of transformative agreements, an evolving
term for contracts that shift library spendings for subscriptions to OA
(Borrego et al., 2020; Hinchliffe, 2019), the demand for
publisher-provided data has even increased. If publishers would enable
transparency about the OA uptake in hybrid journals and APC invoicing,
academic libraries could improve their assessments and adjustments of
contracts with publishers (Antelman, 2017; Schimmer et al., 2015).
Likewise, research organizations and funders could improve compliance
monitoring with OA mandates, while ensuring that the different actors
were not charged multiple times (Larivière \& Sugimoto, 2018). However,
as formal bibliometrics and financial studies have noted, the lack of
publicly available and standardized data about hybrid OA publishing is
considerable (Laakso \& Björk, 2016; Lawson, 2015; Pinfield et al.,
2016)

This current study investigates the extent to which publisher-provided
data can be utilized to research hybrid OA. The study takes Elsevier,
the largest scholarly journal publisher in terms of article volume
(Larivière et al., 2015), as the starting point for an empirical
analysis of hybrid OA uptake and invoicing. Our approach is based on
openly available data sources and brings together scholarly metadata
from Crossref, a DOI registration agency for scholarly works, and
machine-readable APC invoicing data from Elsevier OA full-texts. The
so-compiled data enable us to analyze the development of Elsevier's
hybrid journal portfolio from 2015-19, and to compare it with that of
Elsevier's full and delayed OA program. Further, we investigate whether
hybrid publishing fees were waived or charged. In particular, we
identify academic consortia and research funders that were directly
invoiced by Elsevier, and compare their hybrid OA rates by license and
subject. As such, our findings have implications for research and OA
policy and highlight the potential of publisher-provided data for
large-scale, instantaneous analysis of hybrid OA uptake and invoicing.

@@

The ability to identify and discover hybrid OA articles plays a
fundamental role in monitoring hybrid OA, yet time and again, previous
studies have underscored the difficulties involved in this (cf.
(Akbaritabar \& Stahlschmidt, 2019; Björk, 2012; Emery, 2013; Laakso \&
Björk, 2016; Leeuwen et al., 2018). Insights into the volume of hybrid
OA and the various levels and sources of APC funding could create a
better understanding of the costs involved in scholarly publishing and,
thus, strengthen institutions' bargaining power in negotiations with
publishers, and promote greater equity in scholarly publishing. In the
following section, we examine how previous research has tackled the
challenge of identifying hybrid OA content and what these studies taught
us about the uptake and funding sources of hybrid OA.

@@

In the first article-level study on hybrid OA, Laakso \& Björk (2016)
reported that between 2007 and 2013, the five largest
publishers--Elsevier, Sage, Springer, Taylor \& Francis, and
Wiley--recorded growing numbers of hybrid journals that coincided with a
twenty-fold increase in hybrid OA articles (from 666 to 13,994). As a
result, hybrid journals with at least one OA article more than doubled
in number (1,082 in 2009, 2,714 in 2013) but fell by 31\% relative to
the number of hybrid journals in total. Although Laakso and Björk's
(2016) presented a comprehensive overview tracking the growth of hybrid
OA over time and across multiple publishers, their exploratory approach
is unsuitable for repeated use as it required a considerable amount of
manual data collection and cleaning. Minimizing such manual tasks,
Nelson \& Eggett (2017) narrowed their focus to one publisher--the
American Chemical Society (ACS)--and requested a list of all hybrid OA
articles published from 2006-2011 directly from the ACS. To assess the
hybrid OA uptake, the authors aggregated and compared the number of OA
(n=814) and subscription articles (n=27,621), finding that over the five
years, 2.9\% of ACS articles were published as hybrid OA.

Contrary to the previous two studies, Kirkman (2018) took a research
funder as a starting point, reporting a 26\% share of hybrid OA for
articles funded by the Australian National Health and Medical Research
Council from 2012-2014 (816 of 3,190). Kirkman collected research
articles using the Funding Agency and Funding Text search fields on Web
of Science, and classified articles as OA when they were freely
available on the publisher website, and further distinguished hybrid OA
articles based on journal-level information from the publisher website,
such as hybrid OA journal and APC lists. Another approach is presented
by Pölönen et al. (2020), who used current research information system
(CRIS) data---institutional publication data---to study the extent of OA
among Finnish research publications from 2016-2017 (n=34,507 research
articles). As part of the analysis, the authors identified hybrid OA
articles through a dedicated OA status metadata field that has been
mandatory to report in Finland since 2016 and found a 7\% share of
hybrid OA. However, the drawbacks of Kirkman's (2018) and Pölönen et
al.'s (2020) approaches are that such data sources might not be openly
available or, regarding CRIS data, might not be available at all since
not all countries maintain comprehensive institutional publication data.

Other studies assessed the share of (hybrid) OA for the entire corpus of
scholarly research articles. For instance, using research articles with
an assigned DOI and indexed in the Web of Science Core Collection
(n=2,610,305) as a benchmark, Martín-Martín et al. (2018) assessed how
many research articles from 2009 and 2014 are freely available through
Google Scholar. After identifying OA articles through licensing
information from Crossref, the authors estimated the share of hybrid OA
at around 0.5\% in 2009 and 1.5\% in 2014. In a similar vein, Piwowar et
al. (2018) drew on Crossref to identify content licenses. The authors
followed a data-driven approach based on Unpaywall data, which is openly
available and contained 66 million Crossref DOIs at the time, to
determine how much of the scholarly literature is OA. Based on a random
sample of 100,000 Crossref DOIs, the authors found 3.6\% of hybrid OA,
while pointing towards a growing trend in recent years that recorded a
hybrid OA share of 9.4\% of all articles published in 2015.

Since its launch, Unpaywall has become widely used as an OA evidence
source in bibliometric research and rankings(Huang et al., 2020;
Robinson-Garcia et al., 2020). However, results cannot easily be
compared with each other over time because of Unpaywall's ongoing
changes due to non-standardized data to discover hybrid OA. These
changes mainly reflect the growing albeit incomprehensive license
information in Crossref metadata. To overcome these shortcomings, the
developers extended previous web-scraping approaches to better
distinguish between immediate and delayed OA based on licenses at the
date of observation Piwowar et al. (2019). Relatedly, in October 2020,
Elsevier's license for delayed OA content was added to Unpaywall's list
of license patterns (Unpaywall, n.d.-c). At the journal-level, a more
fine-grained identification of full OA journals using OA article
information was developed to better distinguish them from subscription
journals including hybrid journals (Unpaywall, n.d.-a). With its servie
unsub, Unpaywall recently launched a paid service, where libraries can
compare hybrid OA uptake rates with internal financial and usage data to
assess the value of journal subscriptions (Chawla, 2020).

@@

Some countries like Sweden and the United Kingdom (UK) have negotiated
national publishing agreements that allow authors to publish hybrid OA
at a discounted fee or for free. These agreements can serve as the basis
for studies on the uptake of hybrid OA or cost-benefit analyses (cf.
Olsson (2018); Lawson (2017)) but are often limited in scope as they
tend to focus on a single country and rely on data that seldom is
publicly available. Hence, while being valuable for specific use-cases,
such studies can neither provide a comprehensive understanding of how
the scholarly publishing system functions beyond national boundaries nor
facilitate collective sharing of financial information about hybrid OA.

Pinfield et al.'s (2016) analysis of APC payment records provided by 23
UK higher education institutions revealed a sharp increase in central
payments from 2007-2013, which was largely attributed to the
introduction of block grants by Research Councils UK (RCUK) and
non-compliance sanctions by the Wellcome Trust. Moreover, the study
showed that OA fees were paid almost exclusively through RCUK and
Wellcome block grants (92\%), and only a small number of APCs were paid
through internal funding (7\%). In contrast, a recent Springer Nature
survey found that authors draw on a range of funding sources to cover OA
fees, such as dedicated institutional OA funds, block grants, OA
agreements, or research grants (Monaghan et al., 2020). Most hybrid OA
authors were supported through dedicated institutional OA funds (43\%,
excluding block grants) and OA agreements with Springer Nature (34\%).
The differences between these two studies may be due to changes in
policy and funding arrangements---in particular, the introduction of OA
agreements---since Pinfield et al. (2016), or due to Monaghan et al.'s
(2020) more regionally diverse sample, indicating regional differences
in APC funding.

Regional and policy differences in APC payments also came to light in a
study by Jahn \& Tullney (2016) that analyzed APC records from 30 German
higher education and research institutions crowd-sourced by the Open APC
initiative, the Austrian Science Fund (FWF), Jisc, and the Wellcome
Trust. In particular, the study revealed large differences in the amount
of hybrid OA funded from 2014-2015. Whilst hybrid OA accounted for less
than 1\% of APCs paid by German institutions (23 of 3,846), the three
non-German research funders recorded a hybrid share of75\% (11,533 of
15,779). According to Jahn \& Tullney (2016), this could point towards
differences in science policy, such as hybrid OA being supported by the
three non-German research funders but not by Germany's largest national
funder (DFG). Another possibility is that German hybrid OA fees were
paid from budgets not reported to the Open APC initiative, from where
the authors acquired the German APC data. Among these unreported funds
are research grants and research unit budgets, which author surveys
identified as funding sources for hybrid OA fees (Graaf, 2017; Monaghan
et al., 2020). As such, Jahn and Tullney's (2016) findings could reflect
the complexities and potential limitations of institutional OA spending
data that Pinfield et al.~(2016) and Monaghan et al.~(2020) attributed
to incomplete or missing records.

As this brief review has shown, the most comprehensive studies approach
hybrid OA funding from an institutional perspective that focuses on
research institutions or funders and draws on internal records or
crowd-sourced data like from the Open APC initiative. The advantages of
Open APC data are that they reflect the actual APCs paid rather than
list prices; they are not limited to certain publishers or journals;
they are provided in a standardized format regardless of the publisher.
On the other hand, the drawbacks include the limited number of
participating institutions that are almost exclusively based in Europe
and the dependency on participating organizations for data integrity
(Pieper \& Broschinski, 2018). Moreover, since Open APC data does not
include waivers or free APCs, it cannot present a complete picture of
all publishing activities and expenses. This, however, would pose less
of an issue when analyzing just hybrid OA.

\hypertarget{the-role-of-publisher-provided-data}{%
\subsection{The role of publisher-provided
data}\label{the-role-of-publisher-provided-data}}

Uptake and financial studies have OA policy implications for hybrid OA.
Most prominently, the cOAlition S, an international network of research
funders, released its widely discussed Plan S. In September 2018, the
cOAlition S announced that from 2021 onwards, signatories would no
longer fund articles in hybrid OA journals. This principle includes
so-called ``mirror'' and sister journals--hybrid journals with fully OA
counterparts or split into their fully OA and subscription parts---that
have launched since Plan S was announced but makes an exception for
hybrid journals with transformative agreements (Frantsvåg \& Strømme,
2019; Harrison, 2019; McNutt, 2019; S, n.d.). The Plan S refers to the
Efficiency and Standards for Article Charges (ESAC) guidelines (Geschuhn
\& Stone, 2017) as critical component of transformative agreements,
emphasizing the need for publisher-provided metadata to address the lack
of transparency surrounding hybrid OA. However, studies confirm that
publisher with such agreements lack essential data about hybrid OA
uptake and invoicing, or are not able or willing to provide such
information publicly and instantaneously (Marques et al., 2019; Monaghan
et al., 2020), which, in turn, not just limit the monitoring of hybrid
OA, but also research studies on its size and financial implications.

Here, we focus on Elsevier. The publisher provides article-level
metadata to Crossref, including license information, which can be used
to identify OA articles, and links to HTML and XML full-text versions
for text and data mining purposes. Surprisingly, besides the full-text
article, the XML files also contain invoicing information and the
article's OA status. Table \ref{tab:els_xml} summarizes the key metadata
elements, which comprise the type of entity to which the APC was
invoiced (\texttt{openaccessSponsorType}) and the name of that entity if
available (\texttt{openaccessSponsorName}). Moreover, the article-level
metadata indicates if OA was provided through Elsevier's Open Archive
program (\texttt{openArchiveArticle}), which enables free access to
participating subscription journals after an embargo period (i.e.,
delayed OA; Elsevier (n.d.-b)). As we have been unable to identify
similar efforts by other large publishers (e.g., Springer Nature), this
increased transparency through openly available bibliometric and
financial data at the article level seems unique to Elsevier.

\begin{table}[H]
\caption{\label{tab:els_xml} Overview of hybrid open access (OA) uptake across Elsevier hybrid journals with at least one OA article 2015-2019}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML Node & Description\\
\midrule
\texttt{openaccessArticle} & Indicates if the article is open
access. \\
\texttt{openaccessSponsorType} & Invoice recipient type.\\
\texttt{openaccessSponsorName} & Entity that received Elsevier's APC
invoice.\\
\texttt{openArchiveArticle} & Indicates if open access was provided through
Elsevier's Open Archive programme.\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

In this study, we utilize Elsevier's publicly available metadata to
analyze the uptake and invoicing of their hybrid OA publishing option.
In particular, we use this novel methodology to answer the following
questions:

\begin{itemize}
\item
  What was the uptake of Elsevier's hybrid OA publishing option between
  2015 and 2019?
\item
  Through which channels were hybrid APCs invoiced, and who were the
  recipients?
\end{itemize}

@@

For this study, we collected data relating to Elsevier's hybrid OA
option by drawing on multiple freely available data sources. We
identified Elsevier hybrid journals through Elsevier's APC list and
supplemented our sample with Crossref metadata and text-mined invoicing
data to investigate the financing of immediate OA articles provided
under a Creative Commons license in subscription-based journals. Figure
\ref{fig:workflow} visualizes the automated workflow we used to collect
data from Elsevier and Crossref.

@@

\caption{Data Analytics Workflow to obtain article-level OA and invoicing data using freely available data sources provided by Elsevier and Crossref.
}\label{fig:workflow}
\end{figure}

First, we obtained hybrid journals through an Elsevier APC price list
from May 2020 provided as captured by Matthias et al. (2019). We only
included hybrid journals that were published under the hybrid model
throughout the entire period of our analysis. Hence, we excluded
journals that flipped from OA to hybrid and from hybrid to OA (Matthias,
2020). This approach allowed us to ensure that the article-level
differences we observed were not due changes at the journal-level (i.e.,
increases or decreases in number). Overall, we identified 1,970 unique
hybrid journals.

Next, we used an openly available Crossref database snapshot (Crossref,
2020), which contains all Crossref records registered until March 2020,
to calculate the journals' combined article volume for the five-year
period from 2015-2019. We only included articles published in regular
issues aside from supplements containing conference contributions like
meeting abstracts, indicated by non-numeric pagination. We excluded
non-scholarly journal content, such as the table of contents and lists
of reviewers, following Unpaywall's paratext recognition approach
(Unpaywall, n.d.-b), which we expanded to include patterns indicating
corrections. Furthermore, we categorized the articles by subject
according to the All Science Journal Classification code (ASJC) and
added citation indicators from the Scopus Source Title list (version
June 2020) based on the journal they were published in.

Crossref metadata also contains license information that indicates if
and under which terms the publisher has made articles openly available
on their website (Hendricks et al., 2020). We identified OA articles
through Creative Commons (CC) licenses and then downloaded the XML
version of all CC-licensed articles published in a hybrid journal. From
the XML files, we determined whether or not the articles were immediate
OA using the XML node \texttt{openArchiveArticle} and measured the
uptake of hybrid OA through three bibliometric indicators (Laakso \&
Björk, 2016; Nelson \& Eggett, 2017)-\/---the number of hybrid journals
with at least one OA article, the number of hybrid OA articles, and the
number of hybrid OA articles relative to the number of closed-accessed
articles in hybrid journals with at least one OA article.

Moreover, we obtained APC invoicing metadata from the XML (see Table 1),
which enabled us to identify the invoice channels and recipients of
hybrid OA APCs. Based on the \texttt{openaccessSponsorType} node in the
XML files, we distinguish between four invoice channels, including
invoices billed to the corresponding author, as part of publishing
agreements with funding bodies (hereinafter ``agreements''; cf. Elsevier
(n.d.-a)), fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf.~ref), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of OA agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node.

>>>>

% tightlist
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\title{\textbf{Towards transparency of hybrid open access through publisher-provided metadata: An article-level study of Elsevier}}
\shorttitle{Hybrid OA Elsevier}
\threeauthors{Najko Jahn}{Lisa Matthias}{Mikael Laakso}
\threeaffiliations{Göttingen State and University Library, University of Göttingen\\
Platz der Göttinger Sieben 1, 37073 Göttingen, Germany\\
najko.jahn@sub.uni-goettingen.de
}{Department of Political Science, John F. Kennedy Institute, Freie Universität Berlin \\
Lansstraße 7-9, 14195 Berlin, Germany \\
l.a.matthia@gmail.com
}{Information Systems Science, Hanken School of Economics \\
Arkadiankatu 22, 00100 Helsinki, Finland \\
mikael.laakso@hanken.fi
}

@@

\abstract{With the growth of open access (OA), the financial flows in scholarly journal
publishing have become increasingly complex, but comprehensive data on and
transparency of these flows are still lacking. The opacity is especially
concerning for hybrid OA, where subscription-based journals publish individual
articles as OA if an optional fee is paid. This study addresses the lack of
transparency by leveraging Elsevier article metadata and provides the first publisher-level study of hybrid OA uptake and invoicing. Our
results show that Elsevier's hybrid OA uptake has grown steadily but slowly from
2015-2019, doubling the number of hybrid OA articles published per year and
increasing the share of OA articles in Elsevier's hybrid journals from 2.6\% to
3.7\% of all articles. Further, we find that most hybrid OA articles were
invoiced directly to authors, followed by articles invoiced through agreements
with research funders, institutions, or consortia, with only a few funding
bodies driving hybrid OA uptake. As such, our findings point to the role of
publishing agreements and OA policies in hybrid OA publishing. Our results
further demonstrate the value of publisher-provided metadata to improve the
transparency in scholarly publishing.}

@@

The rise of open access (OA) has added complexity to scholarly
publishing, particularly concerning transparency of economic dimensions.
Financial transparency in journal publishing has long been lacking
because Big Deal subscription contracts between academic institutions
and large academic publishers usually include confidentiality clauses
that prohibit publicly sharing the scope or pricing of such agreements
(Bergstrom et al., 2014; Frazier, 2001; Larivière et al., 2015).
Further, the distributed payment processes to access and publish
scholarly articles add to the opacity of the financial flows of
scholarly publishing (Lawson et al., 2016) as they can involve several
actors simultaneously, including authors, research institutions,
libraries, and funders.

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available by paying an article processing charge (APC),
while the journal as a whole remains subscription-access. Hence, hybrid
OA was originally introduced as a low-risk transitional model that
allows journals to gradually convert to full OA and reduce subscription
costs as the OA uptake increases (D. C. Prosser, 2003). Many large
subscription publishers have since incorporated hybrid OA, but there
have been concerns that the hybrid model allows for double
dipping---receiving two payments for one article, the APC and
subscription fees (D. Prosser, 2015; Shieber, 2009). While publishers
assure they adjust their pricing (Mittermaier, 2015), without
transparency around the uptake of hybrid OA and both revenue streams,
such claims are impossible to evaluate. This is especially concerning
because hybrid OA has recently gained popularity after a decade of only
slow uptake (Björk, 2012). The main drivers of this development have
been research institutions and funders that implemented OA policies,
allocated OA funds, and established formal agreements with publishers
that allow affiliated authors to publish OA free-of-charge (Laakso \&
Björk, 2016).

The recent introduction of transformative agreements, an evolving
concept describing contracts that shift library spending from
subscriptions to OA (Borrego et al., 2020; Hinchliffe, 2019), might
perpetuate the lack of transparency because pricing is often based on
traditional subscription costs. Hence, the demand for publisher-provided
data has increased. More transparency about hybrid OA uptake and funding
could facilitate the assessment and adjustment of publisher contracts
(Schimmer et al., 2015), enhance OA mandate compliance monitoring, and
avoid double-dipping (Larivière \& Sugimoto, 2018). However, previous
studies have noted an absence of transparency in hybrid OA publishing
and a considerable lack of publicly available and standardized data
(Laakso \& Björk, 2016; Lawson, 2015; Pinfield et al., 2016).

This paper investigates to what extent publisher-provided data can
increase the transparency around hybrid OA by taking Elsevier, the
largest scholarly journal publisher (Larivière et al., 2015), as the
starting point for an empirical analysis of hybrid OA uptake and
invoicing. Our approach is based on openly available data sources and
combines scholarly metadata from Crossref, a DOI registration agency,
with APC invoicing data from Elsevier. The compiled data enable us to
analyze the hybrid OA uptake in Elsevier journals between 2015 and 2019
by license and subject, and to compare it with Elsevier's full and
delayed OA program. Further, we determine whether hybrid APCs were
waived or charged to authors or as part of publishing agreements. In the
latter case, we examine the invoiced academic consortia and research
funders. As such, our findings have implications for research and OA
policy and highlight the potential of publisher-provided data for
large-scale, instantaneous analysis of hybrid OA uptake and invoicing.

@@

@@

In the first article-level study on hybrid OA, Laakso \& Björk (2016)
reported that between 2007 and 2013, the five largest
publishers--Elsevier, Sage, Springer, Taylor \& Francis, and
Wiley--recorded growing numbers of hybrid journals that coincided with a
twenty-fold increase in hybrid OA (from 666 to 13,994). As a result,
hybrid journals with at least one OA article more than doubled in number
(1,082 in 2009, 2,714 in 2013) but fell by 31\% relative to the total
number of hybrid journals. Laakso and Björk's (2016) exploratory
approach yielded a comprehensive overview tracking the growth of hybrid
OA over time and across multiple publishers, but due to the amount of
manual data collection and cleaning it is unsuitable for repeated use.
Minimizing such manual tasks, Nelson \& Eggett (2017) narrowed their
focus to one publisher--the American Chemical Society (ACS)--and
requested a list of all hybrid OA articles published from 2006-2011 from
the ACS. To assess the hybrid OA uptake, the authors aggregated and
compared the number of OA (n=814) and subscription articles (n=27,621),
finding that over the five years, 2.9\% of ACS articles were published
as hybrid OA.

Contrary to the previous two studies, Kirkman (2018) took a research
funder as a starting point, reporting a 26\% share of hybrid OA for
articles funded by the Australian National Health and Medical Research
Council from 2012-2014 (816 of 3,190). Kirkman collected research
articles using the Funding Agency and Funding Text search fields on Web
of Science, and classified articles as OA when they were freely
available on the publisher website, and further distinguished hybrid OA
articles based on journal-level information from the publisher website,
such as hybrid OA journal and APC lists. Another approach is presented
by Pölönen et al. (2020), who used current research information system
(CRIS) data---institutional publication data---to study the extent of OA
among Finnish research publications from 2016-2017 (n=34,507 research
articles). As part of the analysis, the authors identified hybrid OA
articles through a dedicated OA status metadata field that has been
mandatory to report in Finland since 2016 and found a 7\% share of
hybrid OA. However, the drawbacks of Kirkman's (2018) and Pölönen et
al.'s (2020) approaches are that such data sources might not be openly
available or, regarding CRIS data, might not be available at all because
not all countries maintain comprehensive institutional publication data.

Other studies assessed the share of (hybrid) OA for the entire corpus of
research articles. Using DOI-assigned research articles indexed in the
Web of Science Core Collection as a benchmark, Martín-Martín et al.
(2018) investigated how many research articles from 2009 and 2014
(n=2,610,305 in total) are freely available through Google Scholar.
After identifying OA articles through licensing information from
Crossref, the authors estimated the share of hybrid OA at around 0.5\%
in 2009 and 1.5\% in 2014. Similarly, in a study assessing the OA level
among all scholarly articles and that experienced by Unpaywall users,
Piwowar et al.~(2018) obtained license information through Crossref and
web-scraping. The study found 3.6\% of 100,000 random Crossref DOIs were
hybrid OA and pointed towards a growing trend in recent years that
recorded a hybrid OA share of 9.4\% of all articles published in 2015.

Since its launch, Unpaywall has become widely used in bibliometric
research and rankings (Huang et al., 2020; Robinson-Garcia et al.,
2020). However, the lack of standardized and comprehensive
publisher-provided data has required several updates to Unpaywall to
improve hybrid OA identification and differentiation from delayed OA
(Piwowar et al., 2019; Unpaywall, n.d.-c, n.d.-a), illustrating ongoing
challenges in tracking and comparing hybrid OA prevalence over time.
After changes in Unpaywall's approach to detect hybrid OA, Piwowar et
al.~(2019) estimated a hybrid OA share of 4\% amongst all articles
published by July 2019.

@@

Pinfield et al.'s (2016) analysis of APC payment records provided by 23
United Kingdom (UK) higher education institutions revealed a sharp
increase in central payments from 2007-2013, which was largely
attributed to the introduction of block grants by Research Councils UK
(RCUK) and non-compliance sanctions by the Wellcome Trust. Moreover, the
study showed that OA fees were paid almost exclusively through block
grants (92\%), and only a small number of APCs were paid through
internal funding (7\%). In contrast, a recent Springer Nature survey
found that authors draw on a range of funding sources to cover OA fees,
such as dedicated institutional OA funds, block grants, OA agreements,
or research grants (Monaghan et al., 2020). Most hybrid OA authors were
supported through dedicated institutional OA funds (43\%, excluding
block grants) and OA agreements with Springer Nature (34\%). The
differences between these two studies might be due to Monaghan et al.'s
(2020) focus on Springer Nature articles or due to\} different policy
and funding arrangements---considering the introduction of OA agreements
since Pinfield et al. (2016) and Monaghan et al.'s (2020) more
regionally diverse sample.

Regional and policy differences in APC payments also came to light in a
study by Jahn \& Tullney (2016) that analyzed APC records from 30 German
higher education and research institutions, the Austrian Science Fund
(FWF), Jisc, and the Wellcome Trust. In particular, the study revealed
large differences in the amount of hybrid OA funded from 2014-2015.
Whilst hybrid OA accounted for less than 1\% of APCs paid by German
institutions (23 of 3,846), the three non-German research funders
recorded a hybrid share of 75\% (11,533 of 15,779). According to Jahn \&
Tullney (2016), this could point towards differences in science policy,
such as hybrid OA being supported by the three non-German research
funders but not by Germany's largest national funder, the German
Research Foundation (DFG). Another possibility is that German hybrid OA
fees were paid from budgets not reported to the Open APC initiative, a
crowd-sourcing effort (Pieper \& Broschinski, 2018) from where the
authors acquired data. Among these unreported funds are research grants
and research unit budgets, which author surveys identified as APC
funding sources (Graaf, 2017; Monaghan et al., 2020). As such, Jahn and
Tullney's (2016) findings could reflect the complexities and potential
limitations of institutional OA spending data that Pinfield et al.
(2016) and Monaghan et al. (2020) attributed to incomplete or missing
records.

In recent years, national consortia in Europe have negotiated publishing
agreements covering hybrid OA fees for affiliated authors (Borrego et
al., 2020). While these have improved invoicing workflows, internal
assessments of hybrid OA uptake and invoicing remain challenging because
transparent and comparative data have largely remained absent (Marques
\& Stone, 2020). Such publicly available information might be scarce
because publishers and institutions lack or withhold such data, or due
to confidentiality clauses (Marques \& Stone, 2020; Monaghan et al.,
2020). Several European consortia have collaborated through the
Efficiency and Standards for Article Charges (ESAC) Initiative and
Knowledge Exchange to standardize metadata requirements across
publishers as great variation has been found between different countries
and publishers (Marques et al., 2019).

\hypertarget{research-questions-and-aims}{%
\subsection{Research questions and
aims}\label{research-questions-and-aims}}

In this paper, we focus on Elsevier, a prominent example in recent
hybrid OA uptake and financial studies (Pinfield et al., 2016; @ Laakso
\& Björk, 2016). Elsevier's OA portfolio presents the challenges in
examining hybrid OA described above. For instance, distinguishing
between different OA types. Elsevier supports delayed (Elsevier,
n.d.-b), hybrid, and full OA, including so-called mirror journals---full
OA counterparts of hybrid journals addressing OA policies opposed to
hybrid OA (Harrison, 2019). Further, Elsevier processes APC invoices
through various channels, such as agreements with research funders and
library consortia or, in the absence thereof, the authors (Elsevier,
n.d.-a, n.d.-c). Surprisingly, we found Elsevier article-level metadata
embedded in XML full-texts indicating the articles' OA status and
invoicing. Leveraging this publicly available data, we address the lack
of transparency around hybrid OA noted in previous studies. In
particular, we use this novel approach to answer the following
questions:

\begin{itemize}
\tightlist
\item
  What was the uptake of Elsevier's hybrid OA publishing option between
  2015 and 2019?
\item
  Through which channels were hybrid APCs invoiced, and who were the
  recipients?
\end{itemize}

@@

For this study, we collected data relating to Elsevier's hybrid OA
option by drawing on multiple freely available data sources. We
identified Elsevier hybrid journals through Elsevier's APC list and
supplemented our sample with Crossref metadata and text-mined invoicing
data to investigate the invoicing of immediate OA articles provided
under a Creative Commons (CC) license in subscription-based journals.
Figure \ref{fig:workflow} visualizes the automated workflow we used to
collect data from Elsevier and Crossref.

@@

\caption{Data Collection Workflow to Obtain Article-Level OA and Invoicing Data.
}\label{fig:workflow}
\end{figure}

First, we identified hybrid journals through an Elsevier APC price list
from May 2020 provided by Matthias (2020). We excluded journals that
transitioned from hybrid to full OA and reverse-flip journals that
flipped from full to hybrid OA (Matthias et al., 2019). Overall, we
identified 1,970 unique hybrid journals that published between 2015 and
2019.

Next, we used an openly available Crossref database snapshot (Crossref,
2020), which contains all Crossref records registered until March 2020,
to calculate the journals' combined article volume for the five-year
period from 2015-2019. We only included articles published in regular
issues aside from supplements containing conference contributions like
meeting abstracts, indicated by non-numeric pagination. We excluded
non-scholarly journal content, such as the table of contents, following
Unpaywall's paratext recognition approach (Unpaywall, n.d.-b), which we
expanded to include patterns indicating corrections. Furthermore, we
categorized the articles by subject according to the All Science Journal
Classification code (ASJC) and added citation indicators from the Scopus
Source Title list (version June 2020) based on the journal they were
published in.

We identified OA articles through CC licenses in Crossref metadata
records (Hendricks et al., 2020) and then downloaded the XML version of
all CC-licensed articles published in a hybrid journal. From the XML
files, we obtained the articles' OA status and invoicing information
(see Table \ref{tab:els_xml}). We determined whether articles were
immediate or delayed OA using the XML node \texttt{openArchiveArticle},
which indicates if an article was made freely available under Elsevier's
Open Archive program, and measured the uptake of hybrid OA. Moreover, we
obtained the invoice channels and recipients of hybrid OA APCs. Using
the \texttt{openaccessSponsorType} node, we distinguished between four
invoice channels, including invoices billed to authors, as part of
publishing agreements with funding bodies (cf. Elsevier (n.d.-a)),
exempted through fee waivers (e.g., in ``cases of genuine need'\,' or
due to society or university sponsorships, cf. Elsevier (n.d.-c)), and
other types not specified by Elsevier. Finally, when hybrid APCs were
invoiced as part of agreements, we identified invoice recipients through
the \texttt{openaccessSponsorName} node. It is worth noting that while
this article focuses on OA articles published in hybrid journals, the
same data is also available for articles published in full OA journals.
Elsevier does not provide the invoiced APC in either case.

\begin{table}[H]
\caption{\label{tab:els_xml}Metadata in Elsevier XML Full-Texts.}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML node & Description\\
\midrule
\texttt{openaccessArticle} & Was the article open
access? \\
\texttt{openaccessSponsorType} & Invoice channel.\\
\texttt{openaccessSponsorName} & Invoice recipient.\\
\texttt{openArchiveArticle} & Was open access provided through
Elsevier's Open Archive program?\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}
